Toward Resolved Simulations of Burning Fronts in Thermonuclear X-ray Bursts

### 2018-09-27 astronum_2018/astronum-xrb.tex
Minor typos

We setup a problem by creating a pair of 1D hydrostatic XRB models (hot and cool)
representing the post- and pre-flame states and apply these laterally
across the domain to setup a hot region at the left end of our domain
in equilibrium with the cool model.  Thermal diffusion and nuclear
reactions quickly setup a laterally propagating flame that begins to
move through our domain.  We resolve the diffusion scale, using a
resolution of 10~cm, and use adaptive mesh refinement (two jumps of
$4\times$) to keep a buffer of low density material at low resolution
above the surface to allow for expansion.  To make the lengthscales
tractable, we rotate the neutron star at 2000~Hz---much faster than
expected.  This helps confine the spreading region to a more
reasonable horizontal region.  Nevertheless, our simulation uses 12288
zones laterally, on the finest grid.  We use the stellar
conducitivities from \cite{Timmes00} and a 13-isotopic alpha-chain
reaction network.  To make the timescales easier, we boost the flame
speed by a factor of 10, by boosting both the conductivity and energy
generation rate.  This is an approximation we hope to relax in the
near future.

Figure~\ref{fig:xrb} shows the flame structure for an initial 2D run
at 0.007~s.  The burning layer is only about 10~m high.  Underneath is
\isot{Ni}{56}, representative of the underlying neutron star, and
above it is very a low density ($10^{-4}~\gcc$) buffer.  Only a
portion of the vertical extent is show.  The top panel shows the
temperature structure.  The initial perturbation reached to $2\times
10^4~\mathrm{cm}$, so at this point, the extent of the burnt region
has nearly doubled.  The flame structure is quite wide, since there
are multiple burning stages beyond just helium to carbon, and this
shows up in the composition.  This is seen more clearly in the next
panel which shows the mean molecular weight of the nuclei ($\bar{A}$).
We see the further behind the head of the flame we are, the heavier
the ash nuclei.  We also see where that the head of the flame
appears lifted off of the base of the burning layer---unburned
material underlies the spreading flame.  This appears
reminiscent of the flame structure seen in the earlier simulations of
\cite{cavecchi:2012}. \MarginPar{Yuri, would you agree?} The third panel shows the energy generation rate.
It is strongest just above the neutron star surface, behind the flame.
We again see that the head of the flame is detached from the base
of the atmosphere, with the strongest burning at the head of the flame
slightly higher up in the atmosphere.   Finally, the bottom
panel shows the velocity through the plane of the simulation---this is
induced by the Coriolis force as the burning front spreads laterally.
A tight hurricane structure has setup as a result of the flame
spreading.  This is the geostrophic balance discussed in
\cite{SPIT_ETAL02}.    These calculations are ongoing and will be
the subject of a detailed study in the near future.

@@

Other ways to make the simulation less expensive can include a subgrid
model, Flame models are commonly used for Type Ia supernovae.  There
however, the flame is thin compared to the scale height, so a simple
model can represent the unresolved flame on the grid.  For the XRB,
the flame thickness is a non-trivial fraction of the pressure scale
height and further, the laminar flame properties will change strongly
over the height of the burning layer (see~\cite{Timmes00}).  This
makes traditional flame models a poor fit.  Methods that use even
something as simple as subzone burning (e.g.~\cite{Wang2012190}) may
help capture the energetics with lower resolution.  This can be
implemented using the AMR machinery.  This may help a little, but will
still be insufficient for full star models.

If we continue to resolve the flame structure, then we need to be
smarter (and more aggressive) about our grid refinement strategy.
Currently, we put the entire fuel layer on the finest grid, but for
the material ahead of the flame, we could keep it at lower resolution
until the flame reaches it.  The main obstacle is that we need to keep
the fuel ahead of the flame in hydrostatic equilibrium, and keeping it
refined accomplishes this well.  Using well-balanced
schemes~\cite{kappeli:2016} can help us retain hydrostatic equilibrium
(HSE) in the lower refined regions, reducing the number of zones
needed to maintain fidelity, but we would still need a way to regrid
the layer ahead of the flame to higher resolution to accurately
capture the burning dynamics.  To accomplish this, we envision
building a refinment strategy that does interpolation of the data onto
the newly refined grids enforcing hydrostatic equilibrium.  This can
allow us to keep only a narrow fully refined region that ``slides''
along the surface of the neutron star keeping pace with the advancing
flame.  Such a scheme would greatly reduce the computational costs.
This is a longer term development.

All of the above developments still involve fully compressible
hydrodynamics, but multiscale methods~\cite{weinan2011principles} where we couple different hydrodynamics
solvers together in a single simulation may ultimately be needed to model the full star while resolving the burning front.
%\MarginPar{Alice and Ian contribute here}
Such techniques rely on it being possible to decompose the dynamics of the system
into two scales, e.g.~a short and a long lengthscale, and that the processes at each
scale are at most weakly coupled with each other. In the context of X-ray bursts, one possibility for
multiscale modelling would be to model the large scale dynamics using a shallow water model
(like that of~\cite{SPIT_ETAL02}) to capture the effects of the Coriolis force, and use
a Low Mach model or
a fully compressible model in the vicinity of the burning front to capture
the turbulent burning processes. Given that the computational costs of the 2d shallow water model
would be negligible compared to those of the compressible model, the overall cost of
this multiscale model would be determined by the size of the compressible region only. This
approach would therefore allow us to capture dynamics across the full star
without sacrificing resolution around the flame.
An example of coupling compressible and Low Mach models within the mesh
refinement framework used by our codes, but in a terrestrial context,
is~\cite{Motheau2018}. An initial implemenation, \cite{Harpole2018}, in an
astrophysical context, coupling shallow water to compressible models in both
Newtonian and relativistic gravity, indicates that considerable computational
efficiencies are possible, but that close attention to the coupling between the
models is needed.

>>>>

We setup a problem by creating a pair of 1D hydrostatic XRB models (hot and cool)
representing the post- and pre-flame states and apply these laterally
across the domain to setup a hot region at the left end of our domain
in equilibrium with the cool model.  Thermal diffusion and nuclear
reactions quickly setup a laterally propagating flame that begins to
move through our domain.  We resolve the diffusion scale, using a
resolution of 10~cm, and use adaptive mesh refinement (two jumps of
$4\times$) to keep a buffer of low density material at low resolution
above the surface to allow for expansion.  To make the lengthscales
tractable, we rotate the neutron star at 2000~Hz---much faster than
expected.  This helps confine the spreading region to a more
reasonable horizontal region.  Nevertheless, our simulation uses 12288
zones laterally, on the finest grid.  We use the stellar
conductivities from \cite{Timmes00} and a 13-isotopic alpha-chain
reaction network.  To make the timescales easier, we boost the flame
speed by a factor of 10, by boosting both the conductivity and energy
generation rate.  This is an approximation we hope to relax in the
near future.

Figure~\ref{fig:xrb} shows the flame structure for an initial 2D run
at 0.007~s.  The burning layer is only about 10~m high.  Underneath is
\isot{Ni}{56}, representative of the underlying neutron star, and
above it is very a low density ($10^{-4}~\gcc$) buffer.  Only a
portion of the vertical extent is shown.  The top panel shows the
temperature structure.  The initial perturbation reached to $2\times
10^4~\mathrm{cm}$, so at this point, the extent of the burnt region
has nearly doubled.  The flame structure is quite wide, since there
are multiple burning stages beyond just helium to carbon, and this
shows up in the composition.  This is seen more clearly in the next
panel which shows the mean molecular weight of the nuclei ($\bar{A}$).
We see that the further behind the head of the flame we are, the heavier
the ash nuclei.  We also see where the head of the flame
appears lifted off of the base of the burning layer---unburned
material underlies the spreading flame.  This appears
reminiscent of the flame structure seen in the earlier simulations of
\cite{cavecchi:2012}. \MarginPar{Yuri, would you agree?} The third panel shows the energy generation rate.
It is strongest just above the neutron star surface, behind the flame.
We again see that the head of the flame is detached from the base
of the atmosphere, with the strongest burning at the head of the flame
slightly higher up in the atmosphere.   Finally, the bottom
panel shows the velocity through the plane of the simulation---this is
induced by the Coriolis force as the burning front spreads laterally.
A tight hurricane structure has setup as a result of the flame
spreading.  This is the geostrophic balance discussed in
\cite{SPIT_ETAL02}.    These calculations are ongoing and will be
the subject of a detailed study in the near future.

@@

Other ways to make the simulation less expensive can include a subgrid
model. Flame models are commonly used for Type Ia supernovae.  There
however, the flame is thin compared to the scale height, so a simple
model can represent the unresolved flame on the grid.  For the XRB,
the flame thickness is a non-trivial fraction of the pressure scale
height and further, the laminar flame properties will change strongly
over the height of the burning layer (see~\cite{Timmes00}).  This
makes traditional flame models a poor fit.  Methods that use even
something as simple as subzone burning (e.g.~\cite{Wang2012190}) may
help capture the energetics with lower resolution.  This can be
implemented using the AMR machinery.  This may help a little, but will
still be insufficient for full star models.

If we continue to resolve the flame structure, then we need to be
smarter (and more aggressive) about our grid refinement strategy.
Currently, we put the entire fuel layer on the finest grid, but for
the material ahead of the flame, we could keep it at lower resolution
until the flame reaches it.  The main obstacle is that we need to keep
the fuel ahead of the flame in hydrostatic equilibrium, and keeping it
refined accomplishes this well.  Using well-balanced
schemes~\cite{kappeli:2016} can help us retain hydrostatic equilibrium
(HSE) in the lower refined regions, reducing the number of zones
needed to maintain fidelity, but we would still need a way to regrid
the layer ahead of the flame to higher resolution to accurately
capture the burning dynamics.  To accomplish this, we envision
building a refinement strategy that does interpolation of the data onto
the newly refined grids enforcing hydrostatic equilibrium.  This can
allow us to keep only a narrow fully refined region that ``slides''
along the surface of the neutron star keeping pace with the advancing
flame.  Such a scheme would greatly reduce the computational costs.
This is a longer term development.

All of the above developments still involve fully compressible
hydrodynamics, but multiscale methods~\cite{weinan2011principles} where we couple different hydrodynamics
solvers together in a single simulation may ultimately be needed to model the full star while resolving the burning front.
%\MarginPar{Alice and Ian contribute here}
Such techniques rely on it being possible to decompose the dynamics of the system
into two scales, e.g.~a short and a long lengthscale, and that the processes at each
scale are at most weakly coupled with each other. In the context of XRBs, one possibility for
multiscale modelling would be to model the large scale dynamics using a shallow water model
(like that of~\cite{SPIT_ETAL02}) to capture the effects of the Coriolis force, and use
a low Mach number or
a fully compressible model in the vicinity of the burning front to capture
the turbulent burning processes. Given that the computational costs of the 2D shallow water model
would be negligible compared to those of the low Mach/compressible model, the overall cost of
this multiscale model would be determined by the size of the low Mach/compressible region only. This
approach would therefore allow us to capture dynamics across the full star
without sacrificing resolution around the flame.
An example of coupling compressible and low Mach number models implemented using the \amrex~framework, but in a terrestrial context,
is~\cite{Motheau2018}. An initial implemenation, \cite{Harpole2018}, in an
astrophysical context, coupling shallow water to compressible models in both
Newtonian and relativistic gravity, indicates that considerable computational
efficiencies are possible, but that close attention to the coupling between the
models is needed.
